State & Transition

### 2020-12-01 StateTransition.Rmd
State and Transition prior

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(raster)
```

@@

```{r}
# cover
cover = raster("conus_land_cover_ARD_tile_h03v04/conus_land_cover_ARD_tile_h03v04.tif")
extent(cover)
nlayers(cover)
res(cover)
crs(cover)
image(cover)

# disturbance
dis.processed = "dis.h03v04.RData"
if(file.exists(dis.processed)){
  load(dis.processed)
} else {
  ## process disturbance data to match cover's extent and resolution
  dis = raster("Cmonster_disturbance_class_ARD_tile_h03v04/Cmonster_disturbance_class_ARD_tile_h03v04.tif")
  extent(dis)
  nlayers(dis)
  res(dis)
  crs(dis)

@@

```{r}
### biomass
#biomass = raster("conus_biomass_ARD_tile_h03v04/conus_biomass_ARD_tile_h03v04.tif")
#extent(biomass)
#res(biomass)
#nlayers(biomass)
#summary(biomass)
```

@@

## Estimate disturbance probability for each cover class
```{r}
fd = freq(dis)

#### disturbance aggregations ####

## 
dis[dis==60] <-

@@

## Determine transitions (cover / pft)

## Reduce dimensionality, save

## select sites of interest, save

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(raster)
options(scipen=999)
```

@@

```{r}
# cover
cover = raster("conus_land_cover_ARD_tile_h03v04/conus_land_cover_ARD_tile_h03v04.tif")
covS  = stack("conus_land_cover_ARD_tile_h03v04/conus_land_cover_ARD_tile_h03v04.tif")
ctime = 1990:2017
extent(cover)
nlayers(cover)
res(cover)
crs(cover)
image(cover)

# disturbance
dis.processed = "dis.h03v04.RData"
if(file.exists(dis.processed)){
  load(dis.processed)
} else {
  ## process disturbance data to match cover's extent and resolution
  #dis = raster("Cmonster_disturbance_class_ARD_tile_h03v04/Cmonster_disturbance_class_ARD_tile_h03v04.tif")
  #extent(dis)
  #nlayers(dis)
  #res(dis)
  #crs(dis)

  disS = stack(x = "Cmonster_disturbance_class_ARD_tile_h03v04/Cmonster_disturbance_class_ARD_tile_h03v04.tif")

@@

```{r}
### biomass
biomass = raster("conus_biomass_ARD_tile_h03v04/conus_biomass_ARD_tile_h03v04.tif")
bioS = stack("conus_biomass_ARD_tile_h03v04/conus_biomass_ARD_tile_h03v04.tif")
#extent(biomass)
#res(biomass)
nlayers(bioS)
#summary(biomass)
```

@@

Disturbance class info from Robert Kennedy (email 2020-11-24)
```
AGENTS
 = {
    1:   'Undisturbed',
    10:  'Unkwn Agent',
    11:  'Other ',
    20:  'Clearcut',
    21:  'Partial Harvest',
    22:  'Salvage',
    30:  'Development',
    31:  'Road',
    40:  'Fire',
    50:  'Insect/Disease',
    51:  'MPB-29',
    52:  'MPB-239',
    53:  'WSB-29',
    54:  'WSB-239',
    61:  'Water',
    62:  'Wind',
    70:  'Avalanche-Chute',
    71:  'Avalanche-Runout',
    72:  'Debris Flow',
    73:  'Landslide',
    100: 'Ukn Slow Disturbance',
    110: 'Ukn Abrupt Disturbance',
    160: 'Recovery',
    200: 'No Vis Change',
    201: 'False Change',
    254: 'Err No Code',
    255: 'Err Undefined'
}

Classes 1 through 50 and 61-73 are based on patches labeled using our methodology (like that 2015 paper you found) to link interpreted patches to agents using random forests.  Unknown agent is a confirmed change that the interpret saw but couldn't attribute.  
51-54 are special cases -- mountain pine beetle and western spruce budworm -- that we took from my former PhD student's work to link ADS data to our time-series stuff.  
Classes 200 and 201 are also based on the interpreter rules for patches -- if they could not determine that there was a change, or if they could determine that the likely change was false -- like a seasonality or snow issue.  Basically both are "no change"

Classes 100-160 are based on pixel-level labels based on the temporal signature alone.  They are smaller than our minimum mapping unit of 11 pixels. 
```

## Estimate disturbance probability for each cover class
```{r}
dtime = 1985:2012
fd = freq(dis)
fdS = freq(disS)
dclass <- unique(unlist(sapply(fdS,function(x){x[,1]}))) ## disturbance classes
dcount = matrix(0,length(dclass),length(fdS))
for(t in 1:length(fdS)){
  sel = match(fdS[[t]][,1],dclass)
  dcount[sel,t] = fdS[[t]][,2]
}
rownames(dcount)=dclass
colnames(dcount)=dtime
d.prior = rowSums(dcount)
d.prior = d.prior/sum(d.prior)
d.prior[d.prior>0.001]   ## classes to worry about

#### disturbance aggregations ####
rcl = matrix(
    ## Untriggered  
       c(1,   0, # Undisturbed
         100, 0, # Ukn Slow Disturbance
         160, 0, # Recovery
         200, 0, # No Vis Change
         201, 0, # False Change
         254, 0, # Err No Code
         255, 0, # Err Undefined
    ## Clearcut
         20, 20, # 'Clearcut',
         21, 20, # Partial Harvest
         22, 20, # Salvage
    ## Fire
         40, 40, # Fire'
    ## Pest
         50, 50, #  Insect/Disease',
         51, 50, # 'MPB-29',
         52, 50, # 'MPB-239',
         53, 50, # 'WSB-29'
         54, 50, # 'WSB-239'
    ## OTHER
         10, 10,  # 'Unkwn Agent',
         11, 10,  # 'Other '
         30, 10, # Development'      ## Development
         31, 10, # Road
         61, 10,  # 'Water',
         62, 10,  # 'Wind',
         70, 10,  # 'Avalanche-Chute',
         71, 10,  # 'Avalanche-Runout',
         72, 10,  # 'Debris Flow',
         73, 10  # 'Landslide',
         ),ncol=2,byrow = TRUE)

disR = raster::subs(disS,as.data.frame(rcl))
save(disR,file=dis.processed)

if(FALSE){
disS[disS==1]   = 0   # Undisturbed
disS[disS==100] = 0   # Ukn Slow Disturbance
disS[disS==160] = 0   # Recovery
disS[disS==200] = 0   # No Vis Change
disS[disS==201] = 0   # False Change
disS[disS==254] = 0   # Err No Code
disS[disS==255] = 0   # Err Undefined

## Cut
##    20:  'Clearcut',
disS[disS==21]  = 20  # Partial Harvest
disS[disS==22]  = 20  # Salvage

## Development
##    30:  'Development',
disS[disS==31]  = 30  # Road

## Fire
##    40:  'Fire',

## Pest
## 50:  'Insect/Disease',
disS[disS==51]  = 50  # 'MPB-29',
disS[disS==52]  = 50  # 'MPB-239',
disS[disS==53]  = 50  # 'WSB-29'
disS[disS==54]  = 50  # 'WSB-239'

## OTHER
##  10:  'Unkwn Agent',
disS[disS==11]  = 10  # 'Other '
disS[disS==61]  = 10  # 'Water',
disS[disS==62]  = 10  # 'Wind',
disS[disS==70]  = 10  # 'Avalanche-Chute',
disS[disS==71]  = 10  # 'Avalanche-Runout',
disS[disS==72]  = 10  # 'Debris Flow',
disS[disS==73]  = 10  # 'Landslide',
}

### RECALCULATE FREQUENCIES

fdS = freq(disR)
save(disR,fdS,file=dis.processed)
dclass <- unique(as.vector(unlist(sapply(fdS,function(x){x[,1]})))) ## disturbance classes
dcount = matrix(0,length(dclass),length(fdS))
for(t in 1:length(fdS)){
  sel = match(fdS[[t]][,1],dclass)
  dcount[sel,t] = fdS[[t]][,2]
}
rownames(dcount)=dclass
colnames(dcount)=dtime
d.prior = rowSums(dcount)
d.prior = d.prior/sum(d.prior)
d.prior[d.prior>0.001]   ## classes to worry about

for(i in seq_along(dclass)){
  plot(dtime,dcount[i,]*(30^2/100^2),main=dclass[i],ylab="hectares",type='l')
}

  ## match disturbance to cover
  disR <- raster::resample(disR,cover,method="ngb")
  dis.resamp = "disR.resamp.RData"
  save(disR,fdS,file=dis.resamp)

@@

## Determine transitions (cover / pft)
## looking at a sample, most disturbances correspond to cover changes THAT YEAR
## exception is fire, where there seems to be a nontrivial number that are forest that year but nonforest e.g barren the following year
## do see a good number of cover transitions
```{r}
train = list()
nd = length(dclass)-1
tmat = matrix(0,10,nd) ## transition matrix
bioBar=bioSq=pbioBar=pbioSq = tmat ## sums for biomass and proportional biomass change
nsamp = 100

## loop over years
ttime = ctime[ctime<2004][-1] ## training years
for(t in seq_along(ttime)){
  print(t)

  if(t>length(train)){ 
## sample
    dtmp <- disR[[which(dtime==ttime[t])]]  ## disturbances for that year
    stmp = raster::sampleStratified(dtmp,size = nsamp,exp=1000) ## sample of cell #'s

## eliminate disturbances that did not apply to forests
    ctmp <- covS[[which(ctime==ttime[t])]]  ## cover for previous year
    ftmp <- stmp[ctmp[stmp[,1]]==42,]       ## only keep conifer
    train[[t]] = list(sample=ftmp)
  } else {
    ftmp = train[[t]]$sample   # restart if already created sample
  }
  save(train,file="stateTrain.RData")

## extract cover and biomass (with restart capability)
  if(is.null(train[[t]][["cover"]])){
    cetmp = raster::extract(covS,ftmp[,1])
    train[[t]]$cover = cetmp
  } else {cetmp = train[[t]][["cover"]]}
  if(is.null(train[[t]][["biomass"]])){
    betmp = raster::extract(bioS,ftmp[,1])
    train[[t]]$biomass = betmp
  } else {betmp = train[[t]][["biomass"]]}
  save(train,file="stateTrain.RData")

  bm0    = apply(betmp[,ctime %in% (ttime[t]-(1:3)),drop=FALSE],1,max)
#  bm1    = betmp[,ctime == ttime[t]]
#  bm2    = betmp[,ctime == (ttime[t]+1)]
#  bm = pmin(bm1,bm2)
  bm  = apply(betmp[,ctime %in% (ttime[t]+(0:2)),drop=FALSE],1,min)
  br    = bm/bm0
  bm[is.na(bm)] = 0
  br[is.na(br)] = 0

  for(c in seq_along(dclass)){
    selc = which(ftmp[,2] == dclass[c])
    plot(ctime,ctime,ylim=range(betmp,na.rm = TRUE),type='n',main=paste(ttime[t],dclass[c]))
    for(i in seq_along(selc)) lines(ctime,betmp[selc[i],],col=floor(cetmp[selc[i],t+1]/10))
    abline(v=ttime[t])
    print(c(t,dclass[c],mean(br[selc])))
  }

## loop over disturbance types
  for(c in seq_along(dclass)){
    print(c(t,c))
    selc = which(ftmp[,2] == dclass[c])

    ## tally within-year transitions
    trans = cetmp[selc,ctime %in% (ttime[t]+0:1),drop=FALSE]
    if(nrow(trans) == 0) next
    trans = floor(trans/10)  ## round down
    trans[trans[,2]==4,1] = 4 ## eliminate 1-yr only transitions
    t1 = unique(trans[,1])
    for(lc in t1){
      if(lc == 4){
        ## for forest, tally next-year transitions
        t2sel = which(trans[,1]==4)
        t2 = unique(trans[t2sel,2])
        tmat[t2,c]    = tmat[t2,c]    + tapply(bm[selc[t2sel]],  trans[,2],length)
        ## for each transition, tally biomass
        bioBar[t2,c]  = bioBar[t2,c]  + tapply(bm[selc[t2sel]],  trans[,2],sum)
        bioSq[t2,c]   = bioSq[t2,c]   + tapply(bm[selc[t2sel]]^2,trans[,2],sum)
        pbioBar[t2,c] = pbioBar[t2,c] + tapply(br[selc[t2sel]],  trans[,2],sum)
        pbioSq[t2,c]  = pbioSq[t2,c]  + tapply(br[selc[t2sel]]^2,trans[,2],sum)
      } else {
        ## tally this-year transitions
        t1sel = which(trans[,1]!=4)
        t0 = unique(trans[t1sel,1])
        tmat[t0,c]    = tmat[t0,c]    + tapply(bm[selc[t1sel]],  trans[,1],length)
        ## for each transition, tally biomass, biomass^2, propBiomass, and propBiomass^2
        bioBar[t0,c]  = bioBar[t0,c]  + tapply(bm[selc[t1sel]],  trans[,1],sum)
        bioSq[t0,c]   = bioSq[t0,c]   + tapply(bm[selc[t1sel]]^2,trans[,1],sum)
        pbioBar[t0,c] = pbioBar[t0,c] + tapply(br[selc[t1sel]],  trans[,1],sum)
        pbioSq[t0,c]  = pbioSq[t0,c]  + tapply(br[selc[t1sel]]^2,trans[,1],sum)
      }

    } ## end loop over lc

  } ## end loop over disturbance class c
} ## end loop over time
TRANS = t(t(tmat)/colSums(tmat))
BBAR  = bioBar/tmat
PBBAR = pbioBar/tmat
BVAR  = sqrt((bioSq -  (bioBar  *  bioBar)/tmat) / (tmat - 1))
PBVAR = sqrt((pbioSq - (pbioBar * pbioBar)/tmat) / (tmat - 1))
BVAR/BBAR - PBVAR
## in all cases, disturbance CV was higher on an absolute than a relative basis
save(PBBAR,PBVAR,file="STprior.RData")
```

## select sites of interest, save
```{r}
## sample disturbances for target year
nsamp = 100
d04 <- disR[[which(dtime==2004)]]
samp = raster::sampleStratified(d04,size = nsamp,exp=1000)

## eliminate disturbances that did not apply to forests
c03 <- covS[[which(ctime==2003)]]
csamp <- c03[samp[,1]]
fsamp <- samp[csamp==42,]

dextract = raster::extract(disR,fsamp[,1])

## plot timeseries for each samp to get a feel for variability and timing
for(c in seq_along(dclass)){
  selc = which(fsamp[,2] == dclass[c])
  plot(dtime,dtime,ylim=range(dextract,na.rm = TRUE),type='n',main=dclass[c])
  for(i in seq_along(selc)) lines(dtime,dextract[selc[i],])
}

## similarly, plot how biomass and cover change
cextract = raster::extract(covS,fsamp[,1])
for(c in seq_along(dclass)){
  selc = which(fsamp[,2] == dclass[c])
  plot(ctime,ctime,ylim=range(cextract,na.rm = TRUE),type='n',main=dclass[c])
  for(i in seq_along(selc)) lines(ctime,jitter(cextract[selc[i],],amount=0.5))
  abline(v=2004)
}

bextract = raster::extract(bioS,fsamp[,1])
for(c in seq_along(dclass)){
  selc = which(fsamp[,2] == dclass[c])
  plot(ctime,ctime,ylim=range(bextract,na.rm = TRUE),type='n',main=dclass[c])
  for(i in seq_along(selc)) lines(ctime,bextract[selc[i],],col=floor(cextract[selc[i],16]/10))
  abline(v=2004)
  print(c(dclass[c],mean(bextract[selc,15]/bextract[selc,14])))
}
